---
title: "PSCI 3075 - Progress report #2 template"
author: "Damon C. Roberts"
format:
    pdf:
        self-contained: true
        colorlinks: true
        documentclass: article
execute:
    echo: false
    warning: false
    message: false
---

```{r}
#| label: setup-block

# Modularly load functions
box::use(
    haven = haven[read_dta],
    dplyr = dplyr[filter, mutate, case_when, select],
    modelsummary = modelsummary[datasummary_skim, modelsummary],
    broom = broom[tidy],
    ggplot2 = ggplot2[ggplot, aes, geom_point, geom_linerange, geom_vline, theme_minimal, labs],
    lindia = lindia[gg_diagnose]
)
# Create dataset list object
data = list()

# Load dataset
data[['Original']] = read_dta("C:/Users/damon/Dropbox/current_projects/white_identity_sources/data/anes-2020/anes_2020_original.dta")
```

# Question 1: What is your research question?

Template: Does X cause Y?

# Question 2: List news *and* academic sources that discuss these two phenomenon and their relationship (there should be more this time than the last time). Provide a summary of what these sources say? Provide a justification as to why your particular analyses tell us something new.

Add more sources than what you had last time. Keep reading.

# Question 3: Describe your theory. What is the structure of the relationship between the two variables of interest? What is the mechanism, or the story, that explains this relationship? Provide a null and alternative hypothesis.

What explains this relationship. Rely on intuition and what you've read.

# Question 4: Define your concepts of interest. How do they differ from similar concepts? Justify why the measures you have from your dataset is an appropriate quantification of your concept.

Don't redo what you did last time. Keep making your discussion of your concepts better. 

# Question 5: Clean the dataset. Provide details about how your variables are coded in your dataset after the cleaning.

```{r}
#| label: cleaning-block

data[['Clean']] = data[['Original']] |>
    #* filtering the dataset to only include white respondents as they are my population of interest and am comparing within this group
    filter(V201549x == 1) |>
    mutate(
        #* Female - Binary variable of gender
            #** coded as: 1 = male, 2 = female
            #** recode to: 0 = male, 1 = female
        Female = case_when(V201600 == 1 ~ 0,
                            V201600 == 2 ~ 1),
        #* White identity - How important is being white to identity
            #** coded as: 1 = extremely important - 5 = not at all important
            #** recode to: 1 = not at all important - 5  = extremely important 
        `White identity` = case_when(V202499x == 1 ~ 5,
                                     V202499x == 2 ~ 4,
                                     V202499x == 3 ~ 3,
                                     V202499x == 4 ~ 2,
                                     V202499x == 5 ~ 1)
    ) |>
    select(Female, `White identity`)
```

Female is coded as 1 if the respondent self-identified as female and 0 if they self-identified as male. BLAH BLAH BLAH

# Question 6: Provide a table of descriptive statistics. Interpret the table and explain what image this gives you of your data.

```{r}
#| label: tbl-bivariate-descriptive-stats
#| tbl-cap: Descriptive statistics

data[['Clean']] |>
    datasummary_skim(histogram = FALSE)
```

@tbl-bivariate-descriptive-stats demonstrates that BLAH BLAH BLAH

# Question 7: Perform a bivariate regression. Provide a table of results and interpret them. What do the results suggest about the relationship between these two variables.

You can either make a table of results like you see in @tbl-bivariate-regression. Or you can make a figure of your results with a coefficient plot like you see in @fig-bivariate-regression. If you present a figure of results, you will want to add an appendix section, and put the table in there. As you saw in Question 6 as well, to reference it, for the label of your codeblock where you generate the table or figure, you add the prefix tbl- for a table or fig- for a figure and name it something. Then you use the @ symbol to then reference the figure or the table. You will also want to add a descriptive caption.

```{r}
#| label: tbl-bivariate-regression
#| tbl-cap: The effect of gender on white political identity

# Create empty model list
model = list()

# Models
model[['Bivariate']] = lm(`White identity` ~ Female, data = data[['Clean']])

# Table of results

model[['Bivariate']] |>
    modelsummary(coef_map = c('Female' = 'Female', '(Intercept)' = 'Constant'), notes = c('Standard errors are in parenthases.', 'Point estimates reflect OLS coefficients.', 'Data source: 2020 American National Election Study.'))

```



```{r}
#| label: fig-bivariate-regression
#| fig-cap: The effect of gender on white political identity

#... already fit and stored model. Can just do a few things to then make coefficient plot out of it.

model[['Bivariate']] |>
    tidy(conf.int = TRUE) |>
    filter(term != '(Intercept)') |>
    ggplot() +
        geom_point(aes(y = term, x = estimate)) +
        geom_linerange(aes(xmin = conf.low, xmax = conf.high, y = term)) +
        geom_vline(aes(xintercept = 0), linetype = 'dashed') +
        theme_minimal() +
        labs(x = 'Estimate', y = '', caption = 'Data source: 2020 American national election study.\nDots represent point estimates from OLS.\n Bars represent 95% confidence intervals.')

```


# Question 8: Perform diagnostics on the bivariate regression you just ran. Interpret what these diagnostics mean for the bias and/or inefficiency in your model? What are the implications for your ability to make claims about the causal relationship between your variables?

```{r, fig.width = 12, fig.height = 12}
#| label: fig-bivariate-diagnoses
#| fig-cap: Diagnoses of bivariate regression model

model[['Bivariate']] |>
    gg_diagnose()
```

Interpret this. Answer the questions in the prompt.


# Question 9: What counfounding variables may be present? Justify why they may threaten your ability to estimate the relationship between your variables of interest. Find the possible confounding variables in your dataset, clean them and provide descriptive statistics for them.

This should be based on theory. What do you know about the world from intuition and from your background reading? What things might effect both concepts?

Do you possibly have alternative predictors that better explain that outcome? Should you include that, why or why not?

Find measures of these concepts that you want to include in a multivariate regression. Clean the data and then provide a table of descriptive statistics by using the datasummary_skim function.

# Question 10: Perform a multivariate regression. Provide a table of results and interpret them. Perform diagnostics on the model. Interpret whether, taken together, your model suggests support for your theory. 

Like you did for questions 7 and 8, run a regression, provide results from the regression, and provide diagnostic plots of the regression. What do the results of the regression imply about your theory? What does the diagnostic plots tell you about the performance of the model?

# Self-evaluation

## Question 1: If you could grade yourself on the effort you put into identifying a research question, the variables, and your data cleaning practices, what would it be? An A, B, C, D, or F?
    
## Question 2: What are you stuck on with your research question?

## Question 3: Anything you would like me to give feedback on?

# Course-evaluation

## Question 1: What can I do to help you in the course?
    
## Question 2: What do you like about the course so far?
    
## Question 3: What do you not like about the course so far?